Datafiles' sources:  

shootings data file: 
https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8

NYC precincts shapefile: 
https://hub.arcgis.com/datasets/DCP::nyc-police-precincts?geometry=-75.284%2C40.524%2C-72.671%2C40.888

NYC Boroughs shapefile:
https://hub.arcgis.com/datasets/498c7ff03780407494301c23cb59b899_0?geometry=-75.284%2C40.524%2C-72.671%2C40.888




Importing all the necessary libraries and the data file

```{r, message = FALSE}     
#Importing all the necessary libraries for the analysis
#install.packages("leaflet")
#install.packages('rgdal')
#install.packages('htmltools') 
library(htmltools)#for adding labels in the maps
library(leaflet) #package used for building maps
library(rgdal) # to read the shape files for boroughs and precincts 
library(tidyverse) #for data manipulation and wrangling
library(ggplot2) #for basic data visualization
library(lubridate) #working with and manipulating dates

datafile <- read_csv("Data/NYPD_Shooting_Incident_Data__Historic_.csv")
View(datafile)

```


Elementary Visualizations from the data file 

```{r, message = FALSE}


data_expanded<- datafile %>% mutate(year_month = format(as.Date(datafile$OCCUR_DATE, 
                                                           '%m/%d/%Y'),'%Y.%m'),
                                    month = format(as.Date(datafile$OCCUR_DATE, 
                                                           '%m/%d/%Y'), '%m'),
                                    year = format(as.Date(datafile$OCCUR_DATE, 
                                       '%m/%d/%Y'), '%Y'), .after = OCCUR_DATE) %>%
                                      rename('Borough' = BORO)
View(data_expanded)


#Visualizing the total number of shootings and murders over the years
shootings_per_year <- data_expanded %>% group_by(year) %>% summarise(n())%>%
                                                rename('num_sh' = `n()`)

murders_per_year <- data_expanded %>% filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  group_by(year) %>% summarise(n()) %>% rename('num_mur' = `n()`)


ggplot(mapping = aes(x= as.double(year)))+
         geom_line(data = shootings_per_year, mapping = aes(y= num_sh), color = 'blue')+ 
         geom_line(data = murders_per_year, mapping = aes(y= num_mur), color = 'red')+ 
  labs(title = "Shooting incidents and consequent murders have decreased over the years", 
       x= 'Year(2006 - 2019)', y = "Number of shootings/murders in New York")



#Visualizing the shooting rate  and murder rate per day over the years
shooting_rate <- shootings_per_year%>% mutate(sh_rate = num_sh/365)
murder_rate <- murders_per_year%>% mutate(mur_rate = num_mur/365)

ggplot(mapping = aes(x= as.double(year)))+
  geom_line(data = shooting_rate, mapping = aes(y = sh_rate), color = 'blue') +
  geom_line(data = murder_rate, mapping = aes(y = mur_rate), color = 'red') +
  labs(title = "Average Shooting rate per day is still over 2.5 while the average murder rate is over 0.5 per day at the end of 2019",
       x= 'Year', y = "Average Shooting/murder rate per day")

```

*Observations*  
1. The number of shootings and associated murders have decreased over the years 
in the time period under consideration.  
2. The overall shooting rate and the corresponding murder rate has also declined
but it still substantial.  


Visualizations to find any obvious pattern in the data in terms of the Boroughs, 
Perpetrators and Victims:

```{r, message = FALSE}

#visualizing the total number of shooting incidents per BORO(by different fills)
ggplot(data = data_expanded, mapping = aes(x= Borough, fill = year))+
  geom_bar()+
  labs(title = "Number of shootings per Borough (2006 - 2019)",
       x= "Borough", y = 'Number of shootings')

ggplot(data = data_expanded, mapping = aes(x= Borough, fill = STATISTICAL_MURDER_FLAG))+
  geom_bar()+
  labs(title = "Number of shootings per Borough (2006 - 2019)",
       x= "Borough", y = 'Number of shootings')+
  geom_text(aes(label=..count..),stat='count',vjust= 0.2)

ggplot(data = data_expanded, mapping = aes(x= Borough, fill = as.character(PRECINCT)))+
  geom_bar()+
  labs(title = "Number of shootings per Borough (2006 - 2019)",
       x= "Borough", y = 'Number of shootings')+
  theme(legend.position = 'none')


#visualizing the Perpetrators' race involved in all shootings 
ggplot(data = data_expanded, mapping = aes(x= PERP_RACE, fill = PERP_RACE))+
  geom_bar()+
  labs(title = "Perpetrator's race involved in the shootings",
       x= "Perpetrator's Race", y = "Number of shootings")+
  theme(axis.text.x = element_text(angle = 90))+
  geom_text(aes(label=..count..),stat='count',vjust=-0.1)


#visualizing the Victims' race involved in all shootings
ggplot(data = data_expanded, mapping = aes(x= VIC_RACE, fill = VIC_RACE))+
  geom_bar()+
  labs(title = "Victims's race involved in the shootings",
       x= "Victim's Race", y = "Number of shootings")+
  theme(axis.text.x = element_text(angle = 90))+
  geom_text(aes(label=..count..),stat='count',vjust=-0.1)



#visualizing the Perpetrators' races involved in all shootings across the different Boroughs
ggplot(data = data_expanded, mapping = aes(x= PERP_RACE, fill = PERP_RACE))+
  geom_bar()+
  facet_wrap(~Borough)+
  labs(title = "Perpetrator's race involved in the shootings across every Borough",
       x= "Perpetrator's Race", y = "Number of shootings")+
  theme(axis.text.x = element_text(angle = 90))+
  geom_text(aes(label=..count..),stat='count',vjust=-0.2)



#visualizing the Victims' race involved in all shootings across the different Boroughs
ggplot(data = data_expanded, mapping = aes(x= VIC_RACE, fill = VIC_RACE))+
  geom_bar()+
  facet_wrap(~Borough)+
  labs(title = "Victims's race involved in the shootings for every Borough",
       x= "Victim's Race", y = "Number of shootings")+
  theme(axis.text.x = element_text(angle = 90))+
  geom_text(aes(label=..count..),stat='count',vjust=-0.2)


#visualizing the Perpetrators' race filled by victims' race
ggplot(data = data_expanded, mapping = aes(x= PERP_RACE, fill = VIC_RACE))+
  geom_bar()+
  labs(title = "Plot of the Perpetrator's race filled by Victim's race",
       x= "Perpetrator's Race", y = "Number of shootings")+
  theme(axis.text.x = element_text(angle = 90))
  
```

*Observations*   
1. Brooklyn and Bronx have seen the maximum number of shootings as well as the
associated number of murders over the years.
2. Blacks and Hispanics have been predominantly involved in the shootings, 
both in terms of the Perpetrators and Victims. This pattern is almost consistent 
through all individual Boroughs.  
3. No clear pattern emerges from the plot of shootings involving Perpetrators against 
their victims.  



Visualizations to find any other pattern in shootings:

```{r, message = FALSE}

shootings_per_day <- datafile%>% group_by(OCCUR_DATE)%>%summarise(n())

#Graph of incidents per day over time
ggplot(data = shootings_per_day, mapping = aes(x= `n()`))+
  geom_histogram(fill = 'blue')+
  labs(title = "The number of incidents seem to follow an exponential decay distribution",
       x= "Number of shootings per day", y = 'Count')

#The graph of incidents per day looks to be exponential

#Trend with hour of the day
time_data <- datafile %>% separate(OCCUR_TIME, into = (c('hr', 'mt', 'sec')), 
                                   sep = ':') %>% group_by(hr) %>% summarise(n())

ggplot(data = time_data, mapping = aes(x= hr, y = `n()`))+
  geom_bar(fill = 'blue', stat = 'identity')+
  labs(title = "Shootings tend to be higher in the night hours",
       x = "Hours of the day", y= 'Number of shootings')


#Trend per month of a year
shootings_per_month <- data_expanded %>% group_by(month)%>%summarise(n())


ggplot(data = shootings_per_month, mapping = aes(x= month, y = `n()`))+
  geom_bar(stat = 'identity', fill = 'blue')+
  labs(title = "The number of shootings are relatively higher during the summer months",
       x = 'Month of a year', y = 'Number of shootings')

#Exploring and visualizing the shooting trend over month/every year
shootings_per_monthyear <- data_expanded %>% group_by(year_month,year) %>% summarise(n()) %>% arrange(year)

ggplot(data = shootings_per_monthyear, mapping = aes(x= year_month, 
                                                     y = `n()`, group = year))+
  geom_point()+
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90))+
  geom_smooth()+
  labs(title = "The number of shootings reaches it's peak in July/August every year",
       x= 'Time(in year/month(2006-2019))', y= 'Number of shootings')



datafile %>% group_by(LOCATION_DESC) %>% summarise(n()) %>% arrange(desc(`n()`))

#Useless visualization
#count_per_day <- data_elaborated %>% arrange(OCCUR_DATE)%>% group_by(OCCUR_DATE) %>% count()
#ggplot(data = count_per_day, mapping = aes(x= OCCUR_DATE, y = n))+
  #geom_point()+
  #theme_minimal()+
  #theme(axis.text.x = element_text(angle = 90))+
  #geom_smooth()

```

*Observations*  
1. The number of shootings seem to follow an exponential decay distribution. More 
on this later in the project. 
2. The number of shootings are relatively higher in the summer months.  
3. The plot of number of shootings per month for each year shows an interesting 
pattern. The distribution over each year seems similar i.e. higher shootings during 
the summer months, further solidifying the case that 
temperature has an affect in the number of shooting incidents.



Interpreting missing data for perpetrators as Unidentified cases:

```{r, message = FALSE}
#Adding a column in the original dataset to identify the cases as solved or unsolved

data_status <- datafile %>% mutate(Perpetrator_status = ifelse(
  is.na(PERP_AGE_GROUP) & is.na(PERP_RACE) &is.na(PERP_SEX),
  'Unidentified', 'Identified'))

data_status  %>% group_by(Perpetrator_status)%>% summarise(n())

perp_identified <- data_status %>% filter(Perpetrator_status == 'Identified')
perp_unidentified <- data_status %>% filter(Perpetrator_status == 'Unidentified')

ggplot(data = data_status, mapping = aes(x= BORO, fill = Perpetrator_status))+
  geom_bar() +
  geom_text(aes(label=..count..),stat='count',vjust= 0.2)
  

```

*Observations*   
1. All the missing observations of the Perpetrator seems to suggest that the Perp 
was not identified. This clearly shows that the case wasn't solved. But at the same 
time, having Perpetrator data only means that the Perp was identified but not 
necessarily captured. 
2. In almost 33% of all the shooting incidents, the perpetrator seem to have got away.  


Working with and exploring maps(static and dynamic) with the leaflet package

```{r, message = FALSE}

#Reading the shape file for the boroughs of New York
ny_counties <- readOGR("Data/NYC_Boroughs-shp/ac61f521-b4ff-4230-87da-ef102865acf2202045-1-1jqmgoq.h2egi.shp")


ny_map <- leaflet() %>%
  addProviderTiles(providers$Stamen.Toner) %>%
  setView(lng = -73.9352, lat = 40.7306,zoom = 11.5) %>%
  addPolygons(data = ny_counties) %>%
addCircleMarkers(lng = data_status$Longitude, lat = data_status$Latitude, 
                 radius = 0.1, color = 'red')
  


leaflet_map <- leaflet() %>% addTiles() %>% 
  setView(lng = -73.9352, lat = 40.7306,zoom = 12) %>%
  addPolygons(data = ny_counties , color = "green", weight = 9, fillColor = 'creme')%>%
addCircleMarkers(lng = perp_identified$Longitude, lat = perp_identified$Latitude, 
                 color = 'green', radius = 0.01, group = 'Perps Identified')%>%
  addCircleMarkers(lng = perp_unidentified$Longitude, lat = perp_unidentified$Latitude, 
                 color = 'red', radius = 0.01, group = 'Perps Unidentified') %>%
  addLayersControl(overlayGroups = c('Perps Identified','Perps Unidentified'),
                   options = layersControlOptions(collapsed = FALSE))

leaflet_map_clustered <- leaflet() %>% 
  addProviderTiles(providers$Stamen.Toner) %>% 
  setView(lng = -73.9352, lat = 40.7306,zoom = 12) %>%
  addPolygons(data = ny_counties , color = "green", weight = 10, fillColor = 'creme')%>%
addCircleMarkers(lng = perp_identified$Longitude, lat = perp_identified$Latitude, 
                 radius = 0.1, color = 'green',group = 'Perps Identified', 
                 clusterOptions = markerClusterOptions())%>%
addCircleMarkers(lng = perp_unidentified$Longitude, lat = perp_unidentified$Latitude, 
                 radius = 0.1, color = 'red', group = 'Perps Unidentified',
                 clusterOptions = markerClusterOptions()) %>%
  addLayersControl(overlayGroups = c('Perps Identified','Perps Unidentified'),
                   options = layersControlOptions(collapsed = FALSE))

leaflet_map

leaflet_map_clustered


#map showing the number of shootings by boroughs

queens_data <- datafile %>% filter(BORO == "QUEENS")
bronx_data <- datafile %>% filter(BORO == "BRONX")
man_data <- datafile %>% filter(BORO == "MANHATTAN")
st_data <- datafile %>% filter(BORO == "STATEN ISLAND")
brook_data <- datafile %>% filter(BORO == "BROOKLYN")


map_boro <- leaflet() %>% addTiles() %>% 
  setView(lng = -73.9352, lat = 40.7306,zoom = 12) %>%
  addPolygons(data = ny_counties , color = "black", weight = 5, fillColor = 'creme')%>%
  
  addCircleMarkers(lng = bronx_data$Longitude, lat = bronx_data$Latitude, 
                 color = 'blue', radius = 0.1, group = 'Bronx')%>%
  addCircleMarkers(lng = brook_data$Longitude, lat = brook_data$Latitude, 
                 color = 'green', radius = 0.1, group = 'Brooklyn') %>%
  addCircleMarkers(lng = man_data$Longitude, lat = man_data$Latitude, 
                 color = 'yellow', radius = 0.1, group = 'Manhattan') %>%
  addCircleMarkers(lng = queens_data$Longitude, lat = queens_data$Latitude, 
                 color = 'red', radius = 0.1, group = 'Queens') %>%
  addCircleMarkers(lng = st_data$Longitude, lat = st_data$Latitude, 
                 color = 'black', radius = 0.1, group = 'Staten Island') %>%
  
  addLayersControl(overlayGroups = c('Bronx','Brooklyn', 'Manhattan', 'Queens', 'Staten Island'),
                   options = layersControlOptions(collapsed = FALSE))

map_boro

```


Chloropleth map with leaflet package:
Choice between chloropleths with Boroughs and Precincts 


```{r, message = FALSE}




ny_precincts <- readOGR("Data/NYC_Police_Precincts/NYC_Police_Precincts.shp")
shootings_pc <- datafile %>% group_by(PRECINCT) %>% summarise(n()) %>%mutate(
  num_shootings = `n()`) %>% mutate(shooting_rate = num_shootings/sum(num_shootings))

shootings_pc <- rbind(shootings_pc, c(22, NA, NA, NA)) 

shootings_pre <- shootings_pc %>% arrange(PRECINCT)

is.element(ny_precincts$Precinct, shootings_pre$PRECINCT)

palette <- colorBin("RdYlBu", domain = shootings_pre$num_shootings)

label_chlor <- paste("<p>", paste('Precinct no.', shootings_pre$PRECINCT),"<p>",
                     "<p>", paste('Num. of shootings', shootings_pre$num_shootings) ,"<p>",
                     sep = '')

ny_shootings_chlor <- leaflet() %>% addProviderTiles(providers$Stamen)%>%
  setView(lng= -74.0060 , lat = 40.7128, zoom = 10)%>%
  addPolygons(data = ny_precincts, color = "Black", weight = 1, smoothFactor = 0.5, 
              highlight = highlightOptions(weight = 3, color = 'grey',
                                           fillOpacity = 0.3,
                                           bringToFront = TRUE),
              fillColor = palette(shootings_pre$num_shootings),
              label = lapply(label_chlor, HTML))%>%
  addLegend(pal = palette, values = shootings_pre$num_shootings, position = "topright")

ny_shootings_chlor


```




